---
title: "Measles Analysis"
format: 
  dashboard:
    nav-buttons:
      - icon: linkedin
        href: www.linkedin.com/in/hailey-hossain
      - icon: envelope
        href: Brandon.mont4@gmail.com
        aria-label: "Email"
    theme: minty
    orientation: columns
embed-resources: true
logo: medical-file.svg
logo-alt: "Medical file logo."
---

```{r}
#| label: packages-setup
#| message: false

library(httr)
library(tidyverse)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)
library(broom)
library(gt)
```

```{r}
#| label: Load Datasets
cases_month <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
covid_cleaned_df <- read.csv("covid_cleaned.csv")
country_gdp <- read.csv("country_gdp.csv")
```


```{r}
#| label: concentration function

calc_concentration <- function(year_in, df = cases_year) {
  df |>
    filter(year == year_in) |>
    mutate(
      global_total = sum(measles_total, na.rm = TRUE),
      share = measles_total / global_total,
      burden_tier = case_when(
        share >= 0.10 ~ "Major (10%+ of global)",
        share >= 0.01 ~ "Significant (1-10%)",
        share >= 0.001 ~ "Minor (0.1-1%)",
        TRUE ~ "Negligible"
      )
    ) |>
    filter(burden_tier != "Negligible") |>
    select(country, measles_total, share, burden_tier) |>
    arrange(desc(share))
}

```

```{r}
#| label: concentration summary
key_years <- cases_year |>
  filter(between(year, 2013, 2023)) |>
  distinct(year) |>
  pull(year)

summarize_concentration <- function(yr) {
  calc_concentration(yr) |>
    summarize(
      year = yr,
      global_total = sum(measles_total),
      top1_country = first(country),
      top1_share = first(share),
      top2_3_share = sum(share[2:3]),
      n_major = sum(burden_tier == "Major (10%+ of global)")
    )
}

concentration_over_time <- map_dfr(key_years, summarize_concentration)
```

```{r}
#| label: rq2 summary 1 function

measles_covid_compare <- function(year1, m_df = cases_year, 
                                  c_df = covid_cleaned_df){
  
  measles_year <- m_df |>
    filter(year == year1) |>
    select(country, measles_total_year = measles_total)
  
  covid_year <- c_df |>
    select(country, starts_with(glue("X{year1}"))) |>
    rowwise() |>
    mutate(covid_total_year = max(c_across(starts_with(glue("X{year1}"))), 
                                  na.rm = TRUE)) |>
    ungroup() |>
    select(country, covid_total_year)
  
  comparison_year <- inner_join(measles_year, covid_year, by = "country") |>
    arrange(desc(covid_total_year)) |>
    slice_head(n = 10) |>
    pivot_longer(cols = c(measles_total_year, covid_total_year), 
                 names_to = "disease", values_to = "cases") |>
    mutate(disease = case_when(disease == glue("measles_total_year") 
                               ~ "Measles", 
                               disease == glue("covid_total_year") 
                               ~ "COVID-19"), 
           country = fct_reorder(country, cases, .fun = max))
  
  return(comparison_year)
  
}
```

```{r}
#| label: rq2 summary 2 function

region_means <- function(year, df1 = cases_month, covid = covid_cleaned_df){
  region_df <- df1 |> 
    distinct(region, country)
  
  mean_df <- covid |> 
    left_join(region_df, by = "country") |> 
    select(country, region, starts_with(glue("X{year}"))) |>
    rowwise() |>
    mutate(covid_total = max(c_across(starts_with(glue("X{year}"))), 
                             na.rm = TRUE)) |>
    ungroup() |>
    select(region, covid_total) |> 
    group_by(region) |> 
    summarise(covid_mean = mean(covid_total)) |> 
    filter(!is.na(region))
  
  return(mean_df)
}
```

```{r}
#| label: join gdp + measles
cases_gdp_join <- inner_join(cases_year, country_gdp, 
                             by = join_by(year, iso3 == country)) |>
    mutate(
      high_measles = measles_lab_confirmed > median(measles_lab_confirmed)
      ) |>
  filter(year == 2019)
```

```{r}
#| label: t test
tt <- t.test(gdp_growth ~ high_measles, data = cases_gdp_join)
```

## Data, RQ's, and RQ1

### Data Context & Research Questions {height="40%"}

::: panel-tabset
## Measles Data

</br>

The main data examined in this study focuses on measles and rubella cases across the world. There are two versions of this data: one that organizes data by month and one that organizes data by year. It was downloaded from the provisional monthly measles and rubella data of the World Health Organization on 2025-06-12. The World Health Organization collected this data from January 2012 to June 2025, and each observation recorded is a country at the specified month or year. The purpose of WHO collecting disease-specific data such as this is to track global outbreaks and aid policy decisions.

## Supplemental Data

</br>

Supplemental data was also used in this study to analyze trends and possible correlations. One of these data sets was focused on hisotrical Covid-19 data from across the world; it contains the total number of Covid-19 cases in each country from January 2020 to March 2023. This data was accessed via a secondary website, called API Ninjas, that collects various sets of data. API Ninjas most likely obtained this data and any country or health-based data from public census data. The second supplemental data set was focused on the Gross Domestic Product (GDP) of each country by year; it was collected from 1980 to 2026, as well as projected up to 2029. This data was also accessed via API Ninjas.

## Research Questions

</br>

**RQ1:** How does the concentration of measles compare across countries?

**RQ2:** Did measles reporting drop during the COVID‑19 pandemic (2020-2021), and does the later rebound suggest it was suppressed?

**RQ3:** Is there a significant difference in mean GDP growth between countries with high vs. low measles burden in 2019?
:::

### RQ1 {height="60%"}
::: {.card title="Research Question 1"}
::: panel-tabset
## Lollipop Graph

</br>

```{r}
#| label: lollipop graph
#| fig-alt: "A lollipop plot showing annual deviation of global measles cases from the 2013–2023 mean, where 2019 stands out dramatically above all other years as the largest single-year surge in the study period."

global_totals <- cases_year |>
  filter(between(year, 2013, 2023)) |>
  group_by(year) |>
  summarize(total_cases = sum(measles_total, na.rm = TRUE)) |>
  mutate(
    mean_cases = mean(total_cases),
    deviation = total_cases - mean_cases,
    above_avg = deviation > 0,
    is_2019 = year == 2019
  )

madagascar_2019_share <- cases_year |>
  filter(year == 2019) |>
  mutate(global_total = sum(measles_total, na.rm = TRUE),
         share = measles_total / global_total) |>
  filter(country == "Madagascar") |>
  pull(share) |>
  scales::percent(accuracy = 0.1)


ggplot(global_totals, aes(x = year, y = deviation)) +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "gray50") +
  geom_segment(aes(xend = year, yend = 0, color = above_avg), 
               linewidth = 1) +
  geom_point(aes(color = above_avg, size = is_2019)) +
  annotate("text", x = 2019,
           y = max(global_totals$deviation) * 0.8,
           label = "2019: largest single-year\nsurge in the study period",
           hjust = 1.15, size = 3.5,
           fontface = "bold", color = "red3") +
  scale_color_manual(
    values = c("TRUE" = "tomato1", "FALSE" = "chartreuse3"),
    labels = c("TRUE" = "Above Average", "FALSE" = "Below Average"),
    name = NULL
  ) +
  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 3), guide = "none") +
  scale_x_continuous(breaks = 2013:2023) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "The 2019 Measles Surge Was an Unprecedented Outlier",
    subtitle = "Annual deviation from the 2013–2023 mean global case count",
    x = NULL,
    y = "Cases Above / Below Average",
    caption = "Figure 1: Visualization 1"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

## Measles Concentration

</br>

```{r}
#| label: concentration table

concentration_over_time |>
  gt() |>
  tab_header(
    title = "Global Measles Concentration by Year",
    subtitle = "How burden is distributed across countries, 2013–2023"
  ) |>
  cols_label(
    year = "Year",
    global_total = "Global Cases",
    top1_country = "Top Country",
    top1_share = "Top Country",
    top2_3_share = "2nd & 3rd Countries",
    n_major = "# Major Burden Countries"
  ) |>
  tab_spanner(
    label = "Share of Global Cases",
    columns = c(top1_share, top2_3_share)
  ) |>
  fmt_percent(
    columns = c(top1_share, top2_3_share),
    decimals = 1
  ) |>
  fmt_number(columns = global_total, use_seps = TRUE, decimals = 0) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "tomato1"),
    locations = cells_body(rows = year == 2019)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = year == 2019)
  ) |>
  tab_footnote(
  footnote = "2019 saw the largest global total in the study period, driven almost entirely by a single outbreak in Madagascar.",
  locations = cells_body(columns = global_total, rows = year == 2019)
) |> 
  tab_caption(
    caption = "Figure 2: Table 1"
  )
```

## Heatmap

</br>

```{r}
#| label: heatmap
#| fig-alt: "A heatmap showing each of the top 15 measles-burden countries' share of global cases from 2013 to 2023, where India shows chronic dark shading across 2013–2017 and Madagascar shows a single dramatic spike in 2019 representing 39.4% of all global cases that year."

top_countries <- cases_year |>
  filter(between(year, 2013, 2023)) |>
  group_by(country) |>
  summarize(total = sum(measles_total, na.rm = TRUE)) |>
  slice_max(total, n = 15) |>
  pull(country)

madagascar_2019_share <- cases_year |>
  filter(year == 2019) |>
  mutate(global_total = sum(measles_total, na.rm = TRUE),
         share = measles_total / global_total) |>
  filter(country == "Madagascar") |>
  pull(share) |>
  scales::percent(accuracy = 0.1)

cases_year |>
  filter(between(year, 2013, 2023)) |>
  group_by(year) |>
  mutate(global_share = measles_total / sum(measles_total, na.rm = TRUE)) |>
  ungroup() |>
  filter(country %in% top_countries) |>
  ggplot(aes(x = year, 
             y = fct_reorder(country, global_share), 
             fill = global_share)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "cornsilk", high = "tomato1",
                      labels = scales::percent,
                      name = "Share of\nGlobal Cases") +
  scale_x_continuous(breaks = 2013:2023) +
  labs(
    title = "Measles Burden Is Dominated by a Handful of Countries",
    subtitle = "Share of global cases among top 15 burden countries, 2013–2023",
    x = NULL, 
    y = NULL,
    caption = "Figure 3: Visualization 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Country Burden

</br>

```{r}
#| label: country_burden
summarize_country_burden <- function(df = cases_year,
                                     start_yr = 2013,
                                     end_yr = 2023) {
  total_global <- df |>
    filter(between(year, start_yr, end_yr)) |>
    pull(measles_total) |>
    sum(na.rm = TRUE)

  df |>
    filter(between(year, start_yr, end_yr)) |>
    group_by(country) |>
    summarize(
      total_cases = sum(measles_total, na.rm = TRUE),
      peak_year = year[which.max(measles_total)],
      peak_cases = max(measles_total, na.rm = TRUE),
      global_share = total_cases / total_global
    ) |>
    arrange(desc(total_cases)) |>
    slice_head(n = 15)
}

summarize_country_burden() |>
  gt() |>
  tab_header(
    title = "Chronic Measles Burden by Country",
    subtitle = "Top 15 countries by total measles cases, 2013–2023"
  ) |>
  tab_caption("Figure 4: Table 2") |>
  cols_label(
    country = "Country",
    total_cases = "Total Cases",
    global_share = "Global Share",
    peak_year = "Peak Year",
    peak_cases = "Peak Cases"
  ) |>
  fmt_number(
    columns = c(total_cases, peak_cases),
    use_seps = TRUE,
    decimals = 0
  ) |>
  fmt_percent(
    columns = global_share,
    decimals = 1
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "tomato1"),
    locations = cells_body(rows = country == "Madagascar")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = country == "Madagascar")
  ) |>
  tab_style(
    style = cell_fill(color = "lightcyan"),
    locations = cells_body(rows = country == "India")
  ) |>
  data_color(
    columns = global_share,
    method = "numeric",
    palette = c("cornsilk", "tomato1")
  ) |>
  tab_footnote(
    footnote = "Madagascar's total is almost entirely from its 2019 outbreak, 
                 unlike India whose burden is spread across multiple years.",
    locations = cells_body(columns = peak_cases, rows = country == "Madagascar")
  )
```

:::
:::

## RQ2 and Model

### RQ2 {height="60%"}

::: {.card title="Research Question 2"}
::: panel-tabset
## Line Graph

</br>

```{r}
#| label: intro visual
#| fig-alt: "Line graph showing the total reported cases of measles across the world from 2019 to 2023. There is a significant drop in number of cases from 2019 to 2020, and the number of cases does not rish again until 2023."

cases_year |>
  filter(between(year, 2019, 2023), country != "Madagascar") |>
  group_by(year) |>
  summarize(total_cases = sum(measles_total, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = total_cases)) +
  geom_line(color = "tomato1", linewidth = 1) +
  geom_point(size = 3, color = "tomato1") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = 2019:2023) +
  labs(
    title = "Global Measles Cases Declined After the 2019 Surge",
    subtitle = "Total reported cases, 2019–2023", 
    caption = "Figure 5: Visualization 3",
    x = NULL, y = "Total Cases"
  ) +
  theme_minimal()
```

## Measles vs. COVID-19

</br>

```{r}
#| label: rq2 summary table1

compare2020 <- measles_covid_compare(2020) |> 
  pivot_wider(names_from = disease, 
              values_from = cases) |> 
  rename("Measles_20" = "Measles", 
         "COVID-19_20" = "COVID-19")
compare2022 <- measles_covid_compare(2022) |> 
  pivot_wider(names_from = disease, 
              values_from = cases) |> 
  rename("Measles_22" = "Measles", 
         "COVID-19_22" = "COVID-19")

inner_join(compare2020, compare2022, by = "country") |> 
  gt() |> 
  fmt_integer(columns = c("Measles_20", "COVID-19_20", 
                          "Measles_22", "COVID-19_22")) |> 
  tab_spanner(
    label = "2020",
    columns = c("Measles_20", "COVID-19_20")
  ) |> 
  tab_spanner(
    label = "2022",
    columns = c("Measles_22", "COVID-19_22")
  ) |> 
  tab_header(
    title = "Total Measles and COVID-19 Cases by Country", 
    subtitle = "Countries with Highest Concentration of COVID-19 Cases in 2020 and 2022"
  ) |> 
  tab_caption(
    caption = "Figure 6: Table 3"
  ) |> 
  cols_label(
    country = "Country", 
    "Measles_20" = "Measles", "Measles_22" = "Measles", 
    "COVID-19_20" = "COVID-19", "COVID-19_22" = "COVID-19"
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## COVID-19 Means

</br>

```{r}
#| label: rq2 summary table2

means_joined <- inner_join(region_means(2020), region_means(2022), 
                           by = "region")

means_joined |> 
  gt() |> 
  fmt_integer(columns = c(covid_mean.x, covid_mean.y)) |> 
  tab_spanner(
    label = "Mean Cases", 
    columns = c(covid_mean.x, covid_mean.y)
  ) |> 
  tab_header(
    title = "Average Number of COVID-19 Cases by Region", 
    subtitle = "Comparison of Mean Cases by Region in 2020 and 2022"
  ) |> 
  tab_caption(
    caption = "Figure 7: Table 4"
  ) |> 
  cols_label(
    region = "Region", 
    covid_mean.x = "2020", covid_mean.y = "2022"
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## Lollipop Graph

</br>

```{r}
#| label: rq2 visualization1
#| fig-alt: "Grouped lollipop chart separated by the different regions. The South-East Asia Region and European Region have clear spikes in mean COVID-19 cases, specifically in 2022."

means_joined |> 
  rename("2020" = covid_mean.x, "2022" = covid_mean.y) |> 
  pivot_longer(cols = c("2020", "2022"), 
               names_to = "year", values_to = "mean") |> 
  ggplot(aes(x = year, y = mean, color = region)) + 
  geom_segment(aes(x = year, xend = year, y = 0, yend = mean), 
               position = position_dodge(width = 0.5)) + 
  geom_point(aes(color = region), size = 4,
             position = position_dodge(width = 0.5)) + 
  scale_color_manual(values = c("aquamarine3", "coral", "darkorchid2", 
                                "deeppink", "darkgreen", "darkgoldenrod1")) +
  facet_wrap(~region) + 
  scale_y_continuous(labels = comma) +
  theme_minimal() + 
  labs(x = "Year", y = "Mean", color = "Region", 
       title = "Average COVID-19 Cases by Region", 
       subtitle = "Comparison Between 2020 and 2022", 
       caption = "Figure 8: Visualization 4")
```

## Bar Graph

</br>

```{r}
#| label: rq2 visualization2
#| fig-alt: "Horizontal bar graph comparing Covid-19 and measles cases in the 10 countries with the highest count of Covid-19 cases. Covid-19 cases significantly outnumber the measles cases."

measles_covid_compare(2020) |> 
  ggplot(aes(y = country, x = cases, fill = disease)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale()), 
                  breaks = 10^(0:8)) +
    scale_fill_manual(values = c("COVID-19" = "#D85A30", 
                                 "Measles" = "#1D9E75")) +
    labs(title    = "Measles vs. COVID-19 Cases in 2020", 
         subtitle = "Top 10 countries by COVID-19 case count — log scale",
         x = "Total Cases (log scale)", y = NULL, fill = NULL, 
         caption = "Figure 9: Visualization 5") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top", panel.grid.major.y = element_blank(), 
          panel.grid.minor = element_blank(), 
          axis.text.y = element_text(size = 10))
```

:::
:::

### Model {height="40%"}
::: {.card title="Research Question 3"}
::: panel-tabset

## Hypotheses
**Null Hypothesis:** There is no difference in mean GPD growth between high and low measles burden countries in 2019

**Alternative Hypothesis:** There is a difference in mean GDP growth between high and low measles burden countries in 2019

## T-Test
```{r}
#| label: "t-test table"

tidy(tt) |>
  mutate(alternative = "Two-sided test (μ₁ ≠ μ₂)") |>
  gt() |>
  tab_header(
    title = md("**Two-Sample t-test Results**"),
    subtitle = "Comparison of mean GDP growth between high and low measles burden countries in 2019"
    ) |>
  cols_label(
    estimate1 = "Mean GDP growth (Low Measles)",
    estimate2 = "Mean GDP growth (High Measles)",
    statistic = "t statistic",
    parameter = "df",
    p.value = "p-value",
    estimate = "Mean difference",
    conf.low = "CI lower",
    conf.high = "CI upper",
    method = "Method",
    alternative = "Alternative Hypothesis"
  ) |>
  fmt_number(columns = everything(), decimals = 4)
```

## Conclusion
Using the 2019 data, a year marked by a notable rise in measles cases, countries were classified as high measles burden if their lab-confirmed measles cases exceeded the median of 48 cases. A two sample t-test resulted in no significant difference in mean GDP growth between countries with high and low measles burden. Due to a t-value = -1.04 and a p-value = 0.299 > alpha = 0.05, we fail to reject the null hypothesis.

:::
:::
